---
title: "Data tidying lesson"
author: "Eugenie Euskirchen"
date: "10/9/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package_load}
library(dplyr)
library(tidyr)
```

To access 'filter' from the 'stats' package:
'status::filter()'

#Read in and clean up data

```{r read in data}
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"),
                    stringsAsFactors = FALSE)
head(catch_original)
```

#Tidy the data
Select the columns we want using the select function, 'select()'
Make sure the pipe is at the end of the line

```{r}
catch_data <- catch_original %>% 
 # select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
  #same as the previous way, but it just says not to take into account rows with All and notesRegCode, instead of just typing the columns we want, as above.
  select(-All, -notesRegCode)
head(catch_data)
```

```{r}
summary(catch_data)
```
Change teh values in the Chinook colum to mumeric using 'mutate()'
```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook=as.numeric(Chinook))

head(catch_clean)
```
Investigate using 'which() and is.no()
Which NAs were introduced by coercion?
```{r}
i <- which(is.na(catch_clean$Chinook))
i
```
Determine which rows have characters in the Chinook column:

```{r}
catch_original[i,]
```

Row 401 has an 'I' instead of a '1'
Now need to fix this in a script, instead of just changing it in the original dataset.
Use 'mutate' to change the 'I' to a '1'
Use 'mutate' to coerce Chinook column to numeric
```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>%
  mutate(Chinook = as.integer(Chinook))

head(catch_clean)
```

Move from wide format to tall format using pivot_longer()
Pivot over all columns except Region and Year
```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to ="species",
               values_to = 'catch')
head(catch_long)
```
Going back to where we were previously using 'catch_wide'
```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = species,
              values_from =catch)
head(catch_wide)
```

Fish units are in thousands
'rename()catch to catch_thousands
```{r}
#catch_long <- catch_long %>% 
 # rename(catch_thousands = catch) %>% 
  #mutate(catch = catch_thousands * 1000) %>% 
  #select(-catch_thousands)
#A better way to do the above
catch_long <- catch_long %>% 
  mutate(catch = catch *1000)

head(catch_long)
```


#Summarize the data
Mean catch by region by species

```{r}
mean_region <- catch_long %>% 
  group_by(Region, species) %>% 
  summarise(mean_catch = mean(catch),
            n_obs = n())
mean_region
```

Mean sockeye catch by region
```{r}
mean_sockeye_region <- catch_long %>% 
  filter(species == "Sockeye") %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch))

mean_sockeye_region
```
Total annual catch
```{r}
annual_catch <- catch_long %>% 
  group_by(Year) %>% 
  summarise(total_catch = sum(catch)) %>% 
  arrange(desc(total_catch))

annual_catch
```
Read in another data table
```{r}
region_defs <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1",
                            method = "libcurl"),
                        stringsAsFactors = FALSE) %>% 
    select(code, mgmtArea)

head(region_defs)
```
Left join the two data sets, specifying with the 'by' statement which columns get joined.
```{r}
catch_joined <-
  left_join(catch_long, region_defs, by =c("Region"= "code"))
head(catch_joined)
```

Another way to do the same thing as above.

```{r}
region_defs <-  region_defs %>% 
  rename(Region = code)

catch_joined <-  left_join(catch_long, region_defs, by = "Region")
catch_joined <-  left_join(catch_long, region_defs)

```
#Seperate and unite example
```{r}
dates_df <- data.frame(date = c("5/24/1930",
                                "5/25/1930",
                                "5/26/1930",
                                "5/27/1930",
                                "5/28/1930"),
                       stringsAsFactors = FALSE)

dates_df %>% 
  separate(date, c("month", "day", "year"), "/", remove = FALSE)

dates_df %>% 
  mutate(three_char = substr(date, start = 1, stop = 3))
```















